Two applications of logic programming to Coq

### 2020-11-07 papers/types2020/elab.tex
Finished a draft of elab.tex.  I'll send email about this section in a few minutes.

%\newpage\quad\newpage
\section{Elaboration of external proof certificates for the Coq kernel}
\label{sec:elab}

@@

\item Some simplifications steps might be applied within a proof
  without recording which rewrites were used.  A simple 
  non-deterministic proof-search engine might be expected to
  reconstruct an equivalent simplification.
\end{enumerate}

As we have describe in Section~\ref{sec:three}, the Foundational Proof
Certificate (FPC) framework has been designed for formally defining a
broad collection of proof certificates that do not necessarily contain
full details.  Proof checkers for FPC certificates can be written
using simple logic programs: the standard implementation mechanisms of
unification and backtracking search can then be used to reconstruct
missing proof details \cite{miller17fac}.  Such proof checking kernels
can also be used to elaborate such certificates into fully detailed
proofs \cite{blanco17phd,blanco17cade}.

\subsection{Moving beyond Horn clauses - because most provers are
  without induction}
\label{ssec:beyond}

If we do not restrict our attention to only Horn clauses then there is
no problem presenting a focused proof system for all of first-order
intuitionistic logic using the connectives $\forall$ and $\supset$.
The following rules extend those given earlier in
Section~\ref{ssec:focused}. 
\[
  \infer[\supset R]{\Pscr\vdash B\supset D}{\Pscr, B\vdash D}
  \qquad
  \infer[\forall R]{\Pscr\vdash \forall x.D}{\Pscr\vdash D[y/x]}
\]
As usual, the $\forall R$ rule has the restriction that $y$ is not
free in its conclusion.

As has been detailed in earlier work on foundational proof
certificates, this richer notion of proof system can provide for
richer proof certificates.  The main differences with what we have
seen before is that the left-hand sides of sequents can now grow
during the proof checking process.  As a result, an indexing mechanism
is used so that it is possible for the decide expert to properly
select locally added assumptions instead of just those assumptions
accepted globally.

To illustrate a proof certificate format that uses this richer logic
and proof system, consider the simple \lP code in
Figure~\ref{fig:debruijn}.  Using the constants provided in that
figure, the untyped $\lambda$-term
$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$ can be
encoded as the following Elpi term of type \lsti{deb}.
%
\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
                                                    [lambda (apply 1 [])])])]))
\end{lstlisting}
Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn
syntax (terms of type \lsti{deb}) are incorporated into proof
certificates (terms of type \lsti{cert}) along with other integer
arguments that are needed to compute offsets to address bound
variables.  
%
This FPC can be interpreted by the proof certificate checker in the
file \verb+ljf-dep.mod+ of the repository
\url{https://github.com/proofcert/fpc-elpi/}.  We note that this proof
certificate checker is also able to simultaneously synthesize a Coq
proof term while checking (see Section~\ref{sec:elab}) all the while
remaining less than 90 lines of Elpi code.

@@

****************

We describe here briefly how to use the technology behind FPCs and
logic programming in order to provide a flexible approach to
connecting external provers of first-order intuitionistic logic to
Coq.
%
Following the general outline that has been described in
\cite{blanco17cade,blanco20coq}, we assume that the following step are
taken.
%
\begin{enumerate}
\item Modify an external prover to output some form of proof evidence
  (proof certificate) for formulas it claims are theorems.

\item Develop a formal definition of the semantics behind such proof
  certificates using the FPC framework.

\item Check proofs by executing the logic programming checker that is
  parameterized by the particular FPC definition.
\end{enumerate}
%
***
The logic programming setting allows such parallel checking
to also serve as checking-and-synthesizing.  That is, during the
checking of one certificate, it is possible to synthesize, for
example, a fully detailed term that is appropriate for handing to the
Coq kernel.

If one is interested only in building Coq kernel proof structures, we
can bypass the use of an explicit pairing operation and build the
synthesis of such proof structures directly into the FPC proof
checker.  (We took exactly this step in Figure~\ref{fig:augmented}
where proof checking involved both proof certificates as well as
Coq-term structures.)  If one is interested in checking only one kind
of external proof structure, then the FPC for that structure could
also be built into the checker (via, say, partial evaluation of logic
programs \cite{lloyd91jlp}).

As an example, consider an external theorem prover for propositional
intuitionistic logic which returns proof structures as untyped
$\lambda$-terms using De Bruijn's notation \cite{}.  For example, the
De Bruijn expression at the end of Section~\ref{ssec:beyond} can be
used as evidence that the following is a theorem of Coq.

@@

Of course, the proof information in the untyped term must be
elaborated to be a fully typed and functional expression.  Proof
checking using the FPC in Figure~\ref{fig:debruijn} and (an extension
of) the proof checker in Figure~\ref{fig:kernel}.

%\newpage

@@

% LocalWords:  intuitionistic dneg peirce forall

>>>>

\section{Elaboration of external proof certificates for the Coq kernel}
\label{sec:elab}

@@

\item Some simplifications steps might be applied within a proof
  without recording which rewrites were used.  A simple 
  non-deterministic proof-search engine might be expected to
  reconstruct an equivalent simplification.
\end{enumerate}
A majority of the external and automatic theorem provers for
intuitionistic logic do not involve induction.  Instead, they go
beyond Horn clause by permitting formulas with no restriction on
occurrences of $\forall$ and $\supset$.  In that case, we need to
modify the focused proof rules that we have seen in
Section~\ref{ssec:focused} by adding the following rules.
\[
  \infer[\supset R]{\Pscr\vdash B\supset D}{\Pscr, B\vdash D}
  \qquad
  \infer[\forall R]{\Pscr\vdash \forall x.D}{\Pscr\vdash D[y/x]}
\]
As usual, the $\forall R$ rule has the restriction that the
eigenvariable $y$ is not free in its conclusion.

As has been detailed in earlier work on foundational proof
certificates, this richer notion of proof system can provide for
richer proof certificates.  The main differences with what we have
seen before is that the left-hand sides of sequents can now grow
during the proof checking process.  When reading the right
introduction rule for $\supset$ from conclusion to premise, we shall
say that the antecedent of the implication is \emph{stored} in the
left side of the context.  When this store action occurs, an
\emph{index} is used by the store command to name that new, left-hand
formula occurrence.  In this extended situation, the decide expert
uses the index of an assumption in order to enter a focus phase of
inference.  A full proof checking kernel for first-order
intuitionistic logic has been given in \cite{chihani17jar} so we do
not reproduce it here.

@@

To give an example, consider using untyped $\lambda$-terms encoded
using De Bruijn's notation as proof certificates for the propositional
intuitionistic logic over just $\supset$.  The fact that such terms
can be used as proof certificates for such formulas (denoting simple
types) can be formally defined using the FPC description in
Figure~\ref{fig:debruijn}.
%
Using the constants provided in that figure, the untyped
$\lambda$-term $\lambda x (x (\lambda y (y (\lambda z (x (\lambda
u\,z))))))$ can be encoded as the following Elpi term of type
\lsti{deb}.
%
\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
                                                    [lambda (apply 1 [])])])]))
\end{lstlisting}
Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn
syntax (terms of type \lsti{deb}) are incorporated into proof
certificates (terms of type \lsti{cert}) along with other integer
arguments that are needed to compute offsets to address bound
variables.

We describe here briefly how to use the technology behind FPCs and
logic programming in order to provide a flexible approach to
connecting external provers of first-order intuitionistic logic to
Coq.  Following the general outline that has been described in
\cite{blanco17cade,blanco20coq}, we assume that the following steps are
taken.
\begin{enumerate}
\item Modify an external prover to output some form of proof evidence
  (proof certificate) for formulas it claims are theorems.

\item Develop a formal definition of the semantics behind such proof
  certificates using the FPC framework.  The FPC for De Bruijn
  expressions given in Figure~\ref{fig:debruijn} is an example of this
  step.

\item Check proofs by executing the logic programming checker that is
  parameterized by the particular FPC definition.
\end{enumerate}
As we have mentioned in Section~\ref{ssec:fpc}, the logic programming
setting allows parallel checking and synthesizing of a pair of
certificates.  That is, during the checking of one certificate, it is
possible to synthesize, for example, a fully detailed term that is
appropriate for handing to the Coq kernel.
If one is interested only in building Coq kernel proof structures, we
can bypass the use of an explicit pairing operation and build the
synthesis of such proof structures directly into the FPC proof
checker.  (We took exactly this step in Figure~\ref{fig:augmented}
where proof checking involved both proof certificates as well as
Coq-term structures.)  If one is interested in checking only one kind
of external proof structure, then the FPC for that structure could
also be built into the checker (via, say, partial evaluation of logic
programs \cite{lloyd91jlp}).

As an example, consider an external theorem prover for propositional
intuitionistic logic which returns proof structures as untyped
$\lambda$-terms using De Bruijn's notation.   Using the FPC provided
in Figure~\ref{fig:debruijn} and the proof certificate checker in the
file \verb+ljf-dep.mod+ of the repository
\url{https://github.com/proofcert/fpc-elpi/},  the De Bruijn term
display above can be elaborated into a proper proof for the following
Coq theorem.

@@

We note that this proof certificate checker and Coq proof synthesizer
is rather compact, comprising less than 90 lines of Elpi code.

@@

% LocalWords:  intuitionistic dneg peirce forall basicstyle lprolog
% LocalWords:  Elpi lc args ljf dep
